How to select predictive models for decision-making or causal inference?

### 2024-10-03 response_to_reviewer.md

# Reviewer 1

I appreciate the approach and the level of detail that this paper offers on an issue that is, to say the least, a hot topic in bioinformatics and data science in general: the evaluation of the best 'error' model when choosing a causal model for predictive purposes. Indeed, there is much confusion on the subject and very often error models are applied without any real awareness of their potential and limitations.

@@

I understand that Eq. (2) and in Eq. (3) consider both treated and non-treated sample (or patient?), but those sums can be also equal to 0, even if their single terms are different from 0. Moreover, it is not clear if, and how Eq. (3) is related to Eq. (2): is it a variant? Is Eq. (3) derived from Eq. (2) under special assumptions?

> Eq. (3) is the definition of the estimator of the ATE based on a model of the outcome. It is sometimes called g-estimation. It  is unbiased under under consistency (Assumption. 3) and unconfoundedness (Assumption 1). To prove that, take the expectation of Eq. (3).  Then apply  Eq. (1) which is true under consistency (Assumption 3) and unconfoundedness (Assumption 1). Then by the law of total expectancy, the conditionnality on X disappear and we recover Eq. (2).
> G-estimation itself does not control error on both populations as highlighted in Figure 1. Specific inductive biases of ML models are able to favor such regularization(see for example [Johannson et al., 2022](https://www.jmlr.org/papers/volume23/19-511/19-511.pdf)). Selection procedures, more elaborated than RMSE are needed to favor such models. I added this last explanation in the method (par. Model-selection risks, oracle and feasible) trying to clarify the link between the hint on error control on both population, g-estimation and the selection procedure which is our main focus.

Furthermore, the Robinson decomposition (R-decomposition) should be better discussed and its role in the context of causal inference should be better commented.
R-decomposition introduces two quantities, the conditional mean outcome and the probability to be treated (known as propensity score). What is the role of the propensity score? How is it calculated? Is it calculated simply by a frequentist approach (number fo treated/total number of patients?

> G-computation is one choice of decomposition of the CATE estimation into a sequence of regressions, where the only statistical estimate is the response function µa(x)  [Chernozhukov et al., 2024, chapter 15](https://causalml-book.org/). Other choices of decomposition exist, such as the R-decomposition, the Double robust estimation, the X-learner... The R-decomposition is one way of identifying the causal quantity of interest which introduce two statistical quantity that have to be modelled: the mean outcome from the treatment effect and the propensity score. Any statistical estimator can be chosen to model these quantities. The frequentist approach is one (naive) such estimator. In practice, researchers often estimate the propensity score with a logistic regression, but other methods such as Boosted trees, Random Forests, or Neural Networks can be used. A good ps estimator should be well calibrated but this question is out of scope for this paper. In our discussion, we briefly mentioned this interesting research question on what properties should have a good estimator for the propensity score.

@@

https://www.research-collection.ethz.ch/handle/20.500.11850/650703

https://aisel.aisnet.org/pacis2022/181/

or some other that can be easily found in literature.

> Thanks for the references. I added two references for Causal ML into the manuscript: [Chernozhukov et al., 2024](https://causalml-book.org/) and [Kaddour et al., 2022](https://arxiv.org/pdf/2206.15475). Both of them cover most of the subfields at the intersection of causal inference and machine learning. I have added them at the beginning of the manuscript before stating our focus on model selection for CATE.

@@

Let me begin with positive comments.  I think the paper is very important because it describes a sophisticated methodology for experimental examination of causal models.
Despite some attempts this is still lacking in literature.  I especially like the simulation scheme which uses different features for data generation and learning models.  Moreover thorough experimental evaluations of causal models are not very frequent making the paper important.  Code is available on Github which is a big plus.

I am however not sure about the overall conclusions, which are that R-risk is always the best.  Performance of risk functions may depend on several factors and the overall best performance might be a result of bias in available data.  The only factor taken into account by the authors is overlap.  However, only one type of lack of overlap is tested in experiments, based on shifted Gaussians.  This will likely lead to violation of the positivity assumption.  How about different scenarios, e.g.: perfect overlap but different treatment/control distributions?  Confounding need not necessarily lead to lack of overlap.  Another possible factor is the ratio of causal effect to background responses.  It is known that it may affect relative performance of HTE estimators.

Overall: I am no fully convinced that R-risk is always the best.  The answer could be dependent on other factors and I hope the authors would explore more scenarios.

> We thank the reviewer for this remark and the suggestion to explore more diverse simulation scenarii. Exploring the sensitivity of our result on the R-risk appears to us particularly necessary. Thus, we ran a dedicated experiment focusing on the ratio of causal effect to background response. We measure the ratio by the absolute mean difference between the causal effect and the baseline response, measured empirically on each dataset instance $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$. We vary a dedicated parameter ($\omega$ in Algorithm 2) in the simulation to force this ratio to cover a wide range of values. In one experiment it ranges from 1 to 300. Figure 7 gives the main results of this experiment. We see that R-risk is still the best risk on average, but when the effect ratio is high (up to 100), the reweighted mu-risk and the the tau-risk_IPW are better than the R-risk. For low value of the ratio, however, the R-risk still outperforms all other metrics. We added these new results to the main paper and the. We also nuanced our conclusion on the R-risk, underlying that it is better on average but that in some situations, other risks might be more adapted. A typical such situation would be when the applied scientist is interested in estimating an effect with a big ratio for the part of the population where there is low overlap.

> We did not explore more sophisticated treatment/control distributions for our simulation. Building these new data generation processes would imply a big modification to our setup. We rely on the easy inversion of the distribution functions of the gaussians (p(X|A=a)) to analytically compute the true propensity score. In addition, we think that the semi-simulated datasets already provide some interesting argument in favor of r-risk with more more diverse treatment/control covariate distributions.

@@

page 17: proposition number is missing

> Thanks for this remark. We fixed it.

A.7: the data generation procedure needs better explanations.  Why not call (7) P(d|A) since that's how its used later?  On page 17 you say '[b1,bD]~P(x)' and later 'x~P(x)' which causes confusion.  Also, how are beta vectors generated?

> We assumed that you talked about the equation 17 rather than 7. Could you point where we use the P(d|A) notation? We did not find it in the manuscript. 
> The notation b1, ..bD ~P(x) was meant to be understood as the beta vectors are drawn from a distribution P(x). The later mention of 'x~P(x)' was meant to be understood as the definition if the RBF function. We removed the second mention which causes confusion.
> The beta vectors are drawn from two normal distributions with mean 0 and unit variances. We added this information in the manuscript.
> We added the Algorithm 2, hopping that it would help the reader to understand exactly how we generate the data for our simulation.

### 2024-10-03 how_to_select_latex.tex
answer to reader, finished v1

\paragraph{The $R\text{-risk}$ is the best metric}

@@

Increasingly complex prediction models integrating richer medical data
have flourished because their predictions can be easily
demonstrated and validated on left-out data. But using them to underpin a
decision on whether to treat or not requires more careful validation,
using a metric accounting for the
putative intervention, the $R\text{-risk}$. The $R\text{-risk}$
brings a sizeable benefit to select the most adequate model, even when
model development is based on treated and
untreated population with little differences, as in RCTs. To facilitate better model selection, we provide Python
code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
% Using the $R\text{-risk}$ does make evaluation more
% complicated not only because the procedure is more involved, but also
% because each intervention requires a dedicated evaluation. However, such
% off-policy evaluation remains much less costly than the recommended good
% practice of impact evaluation testing the ability of a prediction model
% to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
This model-selection procedure puts no constraints on the models used to
build predictive models: it opens the door to evaluating a wide range of
models, from gradient boosting to convolutional neutral, or language
models.

>>>>

\paragraph{The $R\text{-risk}$ is the best metric on average}

@@

Increasingly complex prediction models integrating richer medical data
have flourished because their predictions can be easily
demonstrated and validated on left-out data. But using them to underpin a
decision on whether to treat or not requires more careful validation,
using a metric accounting for the
putative intervention, the $R\text{-risk}$. On average, the $R\text{-risk}$
brings a sizeable benefit to select the most adequate model, even when
model development is based on treated and
untreated population with little differences, as in RCTs. To facilitate better model selection, we provide Python
code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
% Using the $R\text{-risk}$ does make evaluation more
% complicated not only because the procedure is more involved, but also
% because each intervention requires a dedicated evaluation. However, such
% off-policy evaluation remains much less costly than the recommended good
% practice of impact evaluation testing the ability of a prediction model
% to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
This model-selection procedure puts no constraints on the models used to
build predictive models: it opens the door to evaluating a wide range of
models, from gradient boosting to convolutional neutral, or language
models.

\paragraph{Research question needing more work}

On average, the R-risk outperforms other risks, but it is not optimal for every
situations. If one possess some prior knowledge on the data generation process,
another risk might be more adapted. This is typically the case for big causal
effect ratio to baseline and low overlap where the $\tau\text{-risk}_{IPW}$
should be better.
